---
title: "Solving Linear Mixed Models using LMMsolver"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: false
    number_sections: true
bibliography: bibliography.bib
link-citations: yes
vignette: >
  %\VignetteIndexEntry{Solving Linear Mixed Models using LMMsolver}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.dim = c(7, 4)
)
```

```{r setup}
library(LMMsolver)
library(ggplot2)
```

# The LMMsolver package {.unnumbered}

The aim of the `LMMsolver` package is to provide an efficient and flexible system to solve sparse mixed model equations, for models that are often used in statistical genetics. One important application is using splines to model spatial [@Rodriguez-Alvarez2018, @Boer2020] or temporal [@Bustos-Korts2019] trends for phenotypic observations. Another application area is mixed model QTL analysis for multiparental populations, allowing for heterogeneous residual variance and random design matrices with Identity-By-Descent (IBD) probabilities [@Li2021].

The mixed model equations are solved using efficient sparse matrix algebra, using the `spam` package [@Furrer2010]. The automatic differentiation of the Cholesky matrix [@Smith1995] is implemented in C++ using the `Rcpp` package [@Eddelbuettel2018].

# Spatial trend with plots in a single line.

As a first example we will use an oats trial in the `agridat` package. There were 24 varieties in 3 replicates, each consisting of 6 incomplete blocks of 4 plots. The plots were laid out in a single line. For further details of the analysis of this data set using P-splines see @Boer2020.

```{r oatsdata}
## Load data.
data(john.alpha, package = "agridat")
head(john.alpha)
```

In the following mixed model we include rep and gen as fixed effect, and `spl1D` to model a one dimensional P-spline with 20 segments, using the default choice of cubical B-splines and second order differences:

```{r Pspline1D}
## Fit mixed model with fixed and spline part.
obj1 <- LMMsolve(fixed = yield ~ rep + gen,
                 spline = ~spl1D(x = plot, nseg = 20),
                 data = john.alpha)
```

The spatial trend can be obtained as follows:

```{r ObtainSpatialTrend1}
## Extract smooth trend from mixed model.
plotDat <- obtainSmoothTrend(obj1, grid = 1000, includeIntercept = TRUE)
head(plotDat)
```

The trend can then be plotted.

```{r plot1D}
## Plot smooth trend.
ggplot(data = plotDat, aes(x = plot, y = ypred)) +
  geom_line(color = "red", size = 1) +
  labs(title = "Smooth Spatial Trend Oats data", x = "plotnr", y = "yield") +
  theme(panel.grid = element_blank())
```

If first degree B-splines and first order differences are used the P-spline model is equivalent to Linear Variance and Random Walk models, for details see @Boer2020.

```{r Pspline1Db}
obj2 <- LMMsolve(fixed = yield ~ rep + gen,
                 spline = ~spl1D(x = plot, degree = 1, pord = 1, nseg = nrow(john.alpha) - 1),
                 data = john.alpha)
```

The deviance ($dev = -2 \log L_R$) is 

```{r deviance}
round(deviance(obj2), 2)
```
see P-spline(1) model in Table 1 in @Boer2020.

We can obtain a table with effective dimensions (see e.g. @Rodriguez-Alvarez2018) and penalties using the `summary` function:

```{r summaryED}
summary(obj2)
```
The effective dimension is a measure for complexity. In the table above the first three terms are fixed effects and penalized, and therefore the effective dimension is equal to the number of parameters in the model. The term `s(plot)` is a random effect, with effective dimension is \code{14.2}, indicating that there is rather strong non-linear smooth trend, corresponding to what we see in the Figure.  

We can also use the `summary` function to obtain the table with variances:

```{r summaryVAR}
summary(obj2, which = "variances")
```
see Table 1 in @Boer2020. 

# P-splines to fit biomass as function of time.

In this section we show an example of mixed model P-splines to fit biomass as function of time. As an example we take wheat data simulated with crop growth model APSIM. For further details on this simulated data see @Bustos-Korts2019. 

```{r APSIMdat}
data(APSIMdat)
head(APSIMdat)
```
The first column is the environment, Emerald in 1993, the second column simulated genotype (g001), the third column is days after sowing (das), and the last column is the simulated biomass with medium measurement error.

The model can be fitted with
``` {r APSIMmodel}
obj2 <- LMMsolve(biomass ~ 1,
                 spline = ~spl1D(x = das, nseg = 200), 
                 data = APSIMdat)
```

where we took `nseg=200`, just to illustrate that number of B-splines segments cannot be too high (there are only 121 observations), as we can see from the summary table with effective dimension:
``` {r APSIMmodelSummary}
summary(obj2)
```
The fitted smooth trend can be obtained as explained before:

``` {r APSIMplot}
plotDat <- obtainSmoothTrend(obj2, grid = 1000, includeIntercept = TRUE)

ggplot(data = APSIMdat, aes(x = das, y = biomass)) +
  geom_point(size = 1.2) +
  geom_line(data = plotDat, aes(y = ypred), color = "red", size = 1) +
  labs(title = "APSIM biomass as function of time", 
       x = "days after sowing", y = "biomass (kg)") +
  theme(panel.grid = element_blank())
```

The growth rate (first derivative) as function of time can be obtained using `deriv = 1` in function `obtainSmoothTrend`:

```{r APSIMDeriv}
plotDatDt <- obtainSmoothTrend(obj2, grid = 1000, deriv = 1)

ggplot(data = plotDatDt, aes(x = das, y = ypred)) +
  geom_line(color = "red", size = 1) +
  labs(title = "APSIM growth rate as function of time", 
       x = "days after sowing", y = "growth rate (kg/day)") +
  theme(panel.grid = element_blank())
```

# Two-dimensional P-splines.

For two-dimensional mixed P-splines we use the model defined in @Rodriguez-Alvarez2014. As an example we use the `USprecip` data set in the `spam` package, analysed in @Rodriguez-Alvarez2014.

```{r USprecip data}
## Get precipitation data from spam
data(USprecip, package = "spam")

## Only use observed data
USprecip <- as.data.frame(USprecip)
USprecip <- USprecip[USprecip$infill == 1, ]
# 5906 true records, as in 2014 paper.
nrow(USprecip) 
```

The two-dimensional P-spline can be defined with the `spl2D()` function:

```{r runobj3}
obj3 <- LMMsolve(fixed = anomaly ~ 1,
                 spline = ~spl2D(x1 = lon, x2 = lat, nseg = c(41, 41)),
                 data = USprecip)
```

Here the same number of segments is used as in @Rodriguez-Alvarez2014. The summary function gives a table with the effective dimensions and the penalty parameters:

```{r ED_USprecip}
summary(obj3)
```

A plot for the spatial trend can be obtained in a similar way as for the one-dimensional examples:

```{r Plot_USprecip}
plotDat <- obtainSmoothTrend(obj3, grid = c(200, 300), includeIntercept = TRUE)
usa = maps::map("usa", regions = "main", plot = FALSE)
v <- sp::point.in.polygon(plotDat$lon, plotDat$lat, usa$x, usa$y)
plotDat <- plotDat[v == 1, ]

ggplot(plotDat, aes(x = lon, y = lat, fill = ypred)) +
  geom_tile(show.legend = TRUE) +
  scale_fill_gradientn(colors = topo.colors(100))+
  labs(title = "Precipitation (anomaly) US April 1948", x = "Longitude", y = "Latitude") +
  coord_fixed() +
  theme(panel.grid = element_blank())
```

Instead of using the `grid` argument, `newdata` can be used to make predictions for locations specified in a data frame:

```{r newdataARG}
## predictions for new data, using city coordinates from maps package.
data(us.cities, package = "maps")
## Column names have to match column names used for fitting the model.
colnames(us.cities)[colnames(us.cities) == "long"] <- "lon"
head(us.cities)

pred2 <- obtainSmoothTrend(obj3, newdata = us.cities, includeIntercept = TRUE)
head(pred2)
```

# QTL mapping for multiparental populations.

In QTL-mapping for multiparental populations the Identity-By-Descent (IBD) probabilities are used as genetic predictors in the mixed model [@Li2021]. The following simulated example is for illustration. It consists of three parents (A, B, and C), and two crosses AxB, and AxC. AxB is a population of 100 Doubled Haploids (DH), AxC of 80 DHs. The probabilities, pA, pB, and pC, are for a position on the genome close to a simulated QTL:

```{r multipop}
data(multipop)
head(multipop)
```

The residual variances for the two populations can be different. Therefore we need to allow for heterogeneous residual variances, which can be defined by using the `residual` argument in `LMMsolve`:

```{r residualARG}
obj0 <- LMMsolve(fixed = pheno ~ cross, 
                 residual = ~cross, 
                 data = multipop)
logLik(obj0)
```

The QTL-probabilities are defined by the columns pA, pB, pC, and can be included in the random part of the mixed model by using the `group` argument:

```{r groupOPTION}
lGrp <- list(QTL = 3:5)
obj1 <- LMMsolve(fixed = pheno ~ cross, 
                 group = lGrp,
                 random = ~grp(QTL),
                 residual = ~cross,
                 data = multipop) 
logLik(obj1)
coef(obj1)$QTL
```

## References
